Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-07 unimodular-covers.tex
made changes mostly in Section 3

\paco{define mixed IDP, Cayley sums, Minkowski sums, mixed subdivisions? Probably not}
A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding smoothness and IDP for lattice polytopes. The following versions of them are now considered conjectures~\cite{HNPS2008,mfo2007}:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $Q$ refining that of $P$, and $P$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\paco{perhaps we can call this formula ``being mixed IDP''. The property of ``$N(Q)$ refines $N(P)$'' is sometimes called ``$P$ is a Minkowski summand of $Q$''.}

\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\end{enumerate}
\end{conjecture}

Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes. 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds and unimodular simplices, both of which are trivially IDP.

In \Cref{sec:parallelepipeds} we show the following stronger statement:

@@

As a consequence:

@@

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, that we discuss in  \Cref{sec:cayley}. 
As we note there (\Cref{prop:mixedIDP}), if $P,Q\subset \R^d$ are two lattice polytopes one has
\[
\cayley(P,Q) \text{ is IDP} 
\Rightarrow
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\]
In particular, the following statement that we prove in \Cref{sec:cayley} is stronger than the afore-mentioned result of \cite{Fakhruddin,HNPS2008,Ogata}:

@@

This not only implies \Cref{conj:Oda}(1) in dimension three, but also the following two statements, that we prove in \Cref{sec:cayley}.
A $3$-dimensional \emph{prismatoids} is a polytopes whose vertices all lie in two parallel facets. A special case is a polytope of width $1$, whose vertices lie in two consecutive parallel lattice hyperplanes.

@@

\begin{corollary}
Every integer dilation of a lattice $3$-polytope of width $1$ has a unimodular cover.
\end{corollary}

We believe that the $3$-polytopes in all these statements all have unimodular triangulations, but this remains an open question.

@@

We now turn our attention to \emph{prismatoids}, defined in~\cite{Santos-hirsch} as polytopes whose vertices all lie in two parallel facets. Since we are only interested in lattice polytopes, we here use \emph{prismatoid} meaning any $d$-polytope of the form
\[
\conv(Q_1\times\{0\} \cup Q_2 \times \{k\}),
\]
with $k\in \Z_{>0}$ and $Q_1,Q_2$ being lattice $(d-1)$-polytopes. This generalizes Cayley sums, which are the case $k=1$.

\begin{proposition}
\label{prop:prismatoid}
Let $Q_1$ and $Q_2$ be two lattice polygons and consider the prismatoid 
\[
P:= \conv(Q_1\times\{0\} \cup Q_2 \times \{k\},
\]
with $k\ge 2$. 
%Assume that $P\cap(\R^2\times\{i\})$ is a lattice polytope for every $i=0,\dots, k$. This is equivalent to assuming that $P\cap(\R^2\times\{1\})$ is a lattice polytope:
If $P\cap(\R^2\times\{1\})$ is a lattice polygon then $P$ has a unimodular cover.
\end{proposition}

@@

\Cref{coro:prismatoid} follows from \Cref{prop:prismatoid} since smoothness of a prismatoid implies $P\cap(\R^2\times\{i\})$ to be a lattice polytope for every $i$.

@@

\begin{figure}[htb]
\scalebox{.75}{\input{claim2.pdf_t}}
\caption{Illustration of the proof of \Cref{claim:b_and_t}}
\label{fig:claim2}
\end{figure}
\Giulia{fix it to look more unimodular at the bottom}
\begin{figure}[htb]
  \centering
\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
  1]
%coordinate axes
   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%        \tkzLabelX[orig=false,label options={font=\tiny}]
%        \tkzLabelY[orig=false,label options={font=\tiny}]
%        \tkzGrid
%        \tkzDrawX
%        \tkzDrawY

%edges b, t
  \node[
%label={[shift={(0,-0.67)}]$b_1$}, 
mynode] (b1) at (-3,-2) {}; 
  \node[
%label={[shift={(0,-0.67)}]$b_2$}, 
mynode] (b2) at (2, 0.2) {};  
  \node[
%label={[shift={(0,0)}]$t_1$}, 
mynode] (t1) at (-2.7, 1) {};  
  \node[
%label={[shift={(0,0)}]$t_2$}, 
mynode] (t2) at (1.6,0.7) {};

%shaded region where r can be  
  \draw[fill=gray!80, nearly transparent] (2.93,0.61) -- (0,0.81)  -- (0,-0.68) -- cycle;

%p
  \node[
%label={[shift={(0,-0.67)}]$p_1$}, 
mynode] (p1) at (0,-1) {}; 
  \node[
%label={[shift={(0,0)}]$p_2$}, 
mynode] (p2) at (0,1) {};  
%  \node[{p}]  at (-0.3,0.95)

%q
  \node[
%label={[shift={(0,-0.67)}]$q_1$}, 
mynode] (q1) at (-1.7,-0.4) {}; 
  \node[
%label={[shift={(0,-0.67)}]$q_2$}, 
mynode] (q2) at (0.3,-0.4) {};

%segments b, t, p, q
  \draw[ name path = b, black, thick] (b1) node[below, xshift=1.7cm, yshift=0.7cm] {b} -- (b2); 
  \draw[name path = t, black, thick] (t1) node[below, xshift=1.2cm, yshift=-0.1cm] {t} -- (t2);
  \draw[name path = p, black, thick] (p1) node[left, xshift=0 cm, yshift=1.3cm] {p} -- (p2);
  \draw[name path = q, black, thick] (q1) node[below, xshift=0.4cm, yshift=0cm] {q} -- (q2);

%prolonged lines through b and t
 \tkzDrawLine[add=0.2 and 0.3, color=black, dashed](b1, b2);
 \tkzDrawLine[ add=0.3 and 0.4, color=black, dashed](t1, t2);

%functional f_q
  \tkzDefLine[orthogonal=through p1](p1,p2)
  \tkzDrawLine[add = 2 and 1,color=black, dashed](p1,tkzPointResult)
  \node[label={\small $\{f_q(x)=-1\}$}, xshift=0.4cm, yshift=-0.2cm] (lab1) at (3.3, -1) {};

  \tkzDefLine[orthogonal=through p2](p1,p2) 
  \tkzDrawLine[add = 2 and 1,color=black, dashed](p2,tkzPointResult) 
    \node[label={\small $\{f_q(x)=w-1\}$}, xshift=0.4cm, yshift=-0.2cm] (lab2) at (3.3, 1) {};

\end{tikzpicture}
\caption{Picture for Claim 4.2}
\label{fig:claim2}
\end{figure}

@@

 \bibitem{DLRS2010}
J. A. De Loera, J. Rambau, F. Santos
Triangulations: Structures for Algorithms and Applications, 539 pp.
Algorithms and Computation in Mathematics, Vol. 25, Springer-Verlag. 
ISBN: 978-3-642-12970-4

\bibitem{Fakhruddin}
Najmuddin Fakhruddin.
\newblock Multiplication maps of linear systems on smooth projective toric
  surfaces, arXiv:0208.5178, August 2002.

@@

\bibitem{HNPS2008}
Christian Haase, Benjamin Nill, Andreas Paffenholz, and Francisco Santos, Lattice points in Minkowski sums, 
Electron J. Combin., 15 (2008), no. 1, Note 11, 5 pp.

@@

\bibitem{Santos-hirsch}
\paco{add reference}

>>>>

\paco{define mixed IDP, Cayley sums, Minkowski sums, mixed subdivisions in the intro? Probably not}
A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding smoothness and IDP for lattice polytopes. The following versions of them are now considered conjectures~\cite{HNPS2008,mfo2007}:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $P$ refining that of $Q$, and $Q$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\paco{perhaps we can call this formula ``being mixed IDP''. The property of ``$N(Q)$ refines $N(P)$'' is sometimes called ``$P$ is a Minkowski summand of $Q$''.}
\end{enumerate}
\end{conjecture}

Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes. 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

This, together with the mentioned result from~\cite{BHHHJKM2019}, gives:

@@

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, that we discuss in  \Cref{sec:cayley}. 
\paco{added def}
Recall that the \emph{Cayley sum} of two lattice polytopes $Q_1,Q_2\subset \R^d$ is the lattice polytope
\[
\cayley(Q_1,Q_2) := \conv(Q_1\times\{0\} \cup Q_2 \times \{1\}) \subset \R^3.
\]
\paco{new sentence. In fact, this implies that Cayley sums are the same as polytopes of width 1}
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can remove coordinates) but this does not need $Q_1$ and $Q_2$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

As we note there (\Cref{prop:mixedIDP}), if $P,Q\subset \R^d$ are two lattice polytopes one has
\[
\cayley(P,Q) \text{ is IDP} 
\Rightarrow
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\]
In particular, the following statement that we prove in \Cref{sec:cayley} is stronger than the afore-mentioned result of \cite{Fakhruddin,HNPS2008,Ogata}:

@@

This also implies the following two statements, that we prove in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
\paco{Wait; ``width one'' is only a special case of ``prismatoid'' if we allow prismatoids to have vertices in two parallel \emph{faces}, not \emph{facets}. But \Cref{coro:prismatoid} is not so obvious in that more general case. Either prove it or remove the phrase ``special case''}
A special case is a polytope of width $1$, whose vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is equivalent to being $SL(\Z,d)$-equivalent to a Cayley sum.

@@

\begin{corollary}
\label{coro:width1}
Every integer dilation $kP$, $k\ge 2$, of a lattice $3$-polytope $P$ of width $1$ has a unimodular cover.
\end{corollary}

A special case of the latter are integer dilations of empty tetrahedra. That their dilations have
unimodular covers is \cite[Cor.~4.2]{SantosZiegler} (and is also implicit in \cite{KantorSarkaria}).

We believe that the $3$-polytopes in all these statements have unimodular triangulations, but this remains an open question.

@@

We now turn our attention to \emph{prismatoids}, defined in~\cite{Santos-hirsch} as polytopes whose vertices all lie in two parallel facets. In particular, a \emph{lattice prismatoid} is any $d$-polytope $SL(\Z,d)$-equivalent to one of the form
\[
\conv(Q_1\times\{0\} \cup Q_2 \times \{k\}),
\]
where $Q_1,Q_2$ are lattice $(d-1)$-polytopes and $k\in \Z_{>0}$. This generalizes Cayley sums, which are the case $k=1$.

\begin{proposition}
\label{prop:prismatoid}
Let $Q_1$, $Q_2$ be two lattice polygons and consider the prismatoid 
\[
P:= \conv(Q_1\times\{0\} \cup Q_2 \times \{k\},
\]
with $k\ge 2$. 
%Assume that $P\cap(\R^2\times\{i\})$ is a lattice polytope for every $i=0,\dots, k$. This is equivalent to assuming that $P\cap(\R^2\times\{1\})$ is a lattice polytope:
If $P\cap(\R^2\times\{1\})$ is a lattice polygon then $P$ has a unimodular cover.
\end{proposition}

@@

This statement implies \Cref{coro:prismatoid} and \Cref{coro:width1}. 
For the former, smoothness of a prismatoid implies $P\cap(\R^2\times\{1\})$ to be a lattice polytope. \paco{details?}
For the latter, if $P$ has width one then $P\cong \cayley(Q_1, Q_2)$ and, for every $k\in \Z_{>0}$, we have
\[
kP \cap(\R^2\times\{1\}) = (k-1)Q_1 + Q_2.
\]

@@

\begin{figure}[htb]
\scalebox{.75}{\input{claim2.pdf_t}}
\caption{Illustration of the proof of \Cref{claim:b_and_t}}
\label{fig:claim2}
\end{figure}

%\Giulia{fix it to look more unimodular at the bottom}
%\begin{figure}[htb]
%  \centering
%\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
%  1]
%%coordinate axes
%   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%%        \tkzLabelX[orig=false,label options={font=\tiny}]
%%        \tkzLabelY[orig=false,label options={font=\tiny}]
%%        \tkzGrid
%%        \tkzDrawX
%%        \tkzDrawY
%
%%edges b, t
%  \node[
%%label={[shift={(0,-0.67)}]$b_1$}, 
%mynode] (b1) at (-3,-2) {}; 
%  \node[
%%label={[shift={(0,-0.67)}]$b_2$}, 
%mynode] (b2) at (2, 0.2) {};  
%  \node[
%%label={[shift={(0,0)}]$t_1$}, 
%mynode] (t1) at (-2.7, 1) {};  
%  \node[
%%label={[shift={(0,0)}]$t_2$}, 
%mynode] (t2) at (1.6,0.7) {};  
%
%%shaded region where r can be  
%  \draw[fill=gray!80, nearly transparent] (2.93,0.61) -- (0,0.81)  -- (0,-0.68) -- cycle;
%
%
%%p
%  \node[
%%label={[shift={(0,-0.67)}]$p_1$}, 
%mynode] (p1) at (0,-1) {}; 
%  \node[
%%label={[shift={(0,0)}]$p_2$}, 
%mynode] (p2) at (0,1) {};  
%%  \node[{p}]  at (-0.3,0.95)
%
%
%%q
%  \node[
%%label={[shift={(0,-0.67)}]$q_1$}, 
%mynode] (q1) at (-1.7,-0.4) {}; 
%  \node[
%%label={[shift={(0,-0.67)}]$q_2$}, 
%mynode] (q2) at (0.3,-0.4) {};  
%
%%segments b, t, p, q
%  \draw[ name path = b, black, thick] (b1) node[below, xshift=1.7cm, yshift=0.7cm] {b} -- (b2); 
%  \draw[name path = t, black, thick] (t1) node[below, xshift=1.2cm, yshift=-0.1cm] {t} -- (t2);
%  \draw[name path = p, black, thick] (p1) node[left, xshift=0 cm, yshift=1.3cm] {p} -- (p2);
%  \draw[name path = q, black, thick] (q1) node[below, xshift=0.4cm, yshift=0cm] {q} -- (q2);
%
%
%%prolonged lines through b and t
% \tkzDrawLine[add=0.2 and 0.3, color=black, dashed](b1, b2);
% \tkzDrawLine[ add=0.3 and 0.4, color=black, dashed](t1, t2);
% 
%%functional f_q
%  \tkzDefLine[orthogonal=through p1](p1,p2)
%  \tkzDrawLine[add = 2 and 1,color=black, dashed](p1,tkzPointResult)
%  \node[label={\small $\{f_q(x)=-1\}$}, xshift=0.4cm, yshift=-0.2cm] (lab1) at (3.3, -1) {};
%
%
%  \tkzDefLine[orthogonal=through p2](p1,p2) 
%  \tkzDrawLine[add = 2 and 1,color=black, dashed](p2,tkzPointResult) 
%    \node[label={\small $\{f_q(x)=w-1\}$}, xshift=0.4cm, yshift=-0.2cm] (lab2) at (3.3, 1) {};
%
%\end{tikzpicture}
%\caption{Picture for Claim 4.2}
%\label{fig:claim2}
%\end{figure}

@@

 \bibitem{DLRS2010}
J. A. De Loera, J. Rambau, F. Santos
\emph{Triangulations: Structures for Algorithms and Applications}, 539 pp.
Algorithms and Computation in Mathematics, Vol. 25, Springer-Verlag. 
ISBN: 978-3-642-12970-4

\bibitem{Fakhruddin}
Najmuddin Fakhruddin.
\newblock Multiplication maps of linear systems on smooth projective toric surfaces.
  Preprint, arXiv:0208.5178, August 2002.

@@

\bibitem{HNPS2008}
Christian Haase, Benjamin Nill, Andreas Paffenholz, and Francisco Santos, Lattice points in Minkowski sums, 
\emph{Electron.~J.~Combin.}, 15 (2008), no. 1, Note 11, 5 pp.

\bibitem{KantorSarkaria}
Jean-Michel Kantor and Karanbir~S.\ Sarkaria.
On primitive subdivisions of an elementary tetrahedron, 
\emph{Pacific J.~Math.} \textbf{211} (2003), no.~1,  123--155.

@@

\bibitem{Santos-hirsch}
Francisco Santos.
A counter-example to the Hirsch Conjecture.
\emph{Ann.~Math.~(2)}, 176 (July 2012), 383--412. 
DOI: 10.4007/annals.2012.176.1.7

\bibitem{SantosZiegler}
Francisco Santos and G{\"u}nter~M. Ziegler.
\newblock {Unimodular triangulations of dilated 3-polytopes}, 
\newblock {\em Trans. Moscow Math. Soc.} (2013), 293--311. % \MR{3235802}
